---
title: "RTT Improvement Journey"
format:
  html:
    toc: true
    embed-resources: true
params:
  referral_rates: null
---

```{r}
#| echo: false
library(NHSRwaitinglist)
knitr::opts_chunk$set(fig.path = "rtt_improvement_report_files/figure-pdf/")
```

## Introduction

This report summarises key waiting list information for services of different sizes.
Services of different sizes have patients who arrive at different rates, often called the incoming referral rate. (e.g. 250 new referrals per week)

Given a referral rate, operations management and queuing theory mathematics can be used to calculate the relationship between the number of patients waiting, and the RTT performance (referral to treatment time). RTT performance is linked to the number of patients waiting, and cannot be sustainably influenced by any mechanism other than changing the waiting list size.  

In turn, the only mechanism that can influence the waiting list size is the ratio between service demand (referral rate), and service capacity (treatment rate or removal rate). This ratio is the "load", and the load must be less than 1 for waiting lists to be sustainably managed.  

More details are explored and brought to life in the video and white paper linked in the [further information](#further-information) section below.  

```{r}
#| echo: false
#| results: asis
#| fig-width: 8
#| fig-height: 5
#| out-width: "100%"
#| message: false
#| warning: false

rr <- params$referral_rates

# map over each of the referral rates to produce a plot for each
purrr::walk(rr, function(x){
  
  cat("---")
  cat("\n\n")
  cat("## ", x, " Referrals per week \n\n")
  print(plot_rtt_journey(x))
  cat("\n\n")
  
})

```


## Further information {#further-information}

This report has been generated by the {NHSRwaitinglist} R package, an open-source analytical software development project created jointly by the NHSR Community and Professor Neil Walton of Durham University Business School.

The report was originally created for the package by Tom Smith, Head of Activity Analysis & Forecasting at Nottingham University Hospitals NHS Trust.  

The code is MIT-licenced and available for review (and contributions) here:  
<https://github.com/nhs-r-community/NHSRwaitinglist>

There is a methodology overview video here:  
<https://www.youtube.com/watch?v=NWthhW5Fgls>

There is a white paper here:  
<https://www.medrxiv.org/content/10.1101/2022.08.23.22279117v1.full>

END