---
title: "Analysing Waiting Lists with Histogram Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analysing with histogram data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.height = 5,
  fig.width = 8
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup}
library(NHSRwaitinglist)
library(dplyr)
library(ggplot2)

# set up a theme for our charts
theme_set(
  theme_minimal() +
    theme(
      plot.subtitle = element_text(face = "italic")
    )
)
```

# Introduction

The NHSRwaitinglist package provides two complementary approaches for analysing waiting lists:

1. **Patient-level data**: Individual records with referral and removal dates
2. **Histogram data**: Aggregated counts by waiting time intervals

This vignette focuses on the **histogram approach**, which is particularly valuable for:

- **Processing NHS national statistics**: Published data is aggregated, not patient-level
- **Privacy-preserving analysis**: No individual patient records required
- **Large dataset performance**: Efficient computation with time-bucketed summaries
- **Historical data**: Working with pre-aggregated reported statistics

## When to Use Histogram Functions

Use histogram functions when your data is:

- Aggregated into time bins (weeks, months) with patient counts
- Published in summary form (e.g., NHS England RTT statistics)
- Too large for patient-level processing
- Subject to privacy constraints

## Histogram Data Structure

A histogram in this package is a data frame with these essential columns:

- `arrival_since` (Date): Start of the arrival time interval
- `arrival_before` (Date): End of the arrival time interval  
- `n` (Numeric): Count of patients in that interval
- `report_date` (Date): Reference date of the waiting list snapshot

Additional categorical columns (e.g., specialty, priority) are optional.

# Creating Histogram Data

## Example: Converting Weekly Bins

```{r create-histogram}
# Example: Waiting list snapshot from 31 Jan 2024
# Bins represent weeks waiting
wl_hist_jan <- data.frame(
  arrival_since = as.Date(c("2024-01-24", "2024-01-17", "2024-01-10", "2024-01-03")),
  arrival_before = as.Date(c("2024-01-30", "2024-01-23", "2024-01-16", "2024-01-09")),
  n = c(45, 38, 52, 61),  # Patients waiting 0-1, 1-2, 2-3, 3-4 weeks
  report_date = as.Date("2024-01-31")
)

print(wl_hist_jan)
```

The first row shows 45 patients who arrived between 24-30 Jan (waiting < 1 week on 31 Jan).

## Multiple Time Points

For removal statistics, we need multiple snapshots:

```{r multiple-snapshots}
# Add February snapshot
wl_hist_feb <- data.frame(
  arrival_since = as.Date(c("2024-02-21", "2024-02-14", "2024-02-07", "2024-01-31")),
  arrival_before = as.Date(c("2024-02-27", "2024-02-20", "2024-02-13", "2024-02-06")),
  n = c(42, 35, 48, 55),
  report_date = as.Date("2024-02-29")
)

# Combine into single histogram
wl_hist <- rbind(wl_hist_jan, wl_hist_feb)

# View unique report dates
unique(wl_hist$report_date)
```

## Using format_histogram()

The `format_histogram()` function ensures your data has the correct structure:

```{r format-histogram}
# If you only have arrival_since, format_histogram adds arrival_before
hist_simple <- data.frame(
  arrival_since = as.Date(c("2024-01-24", "2024-01-17", "2024-01-10")),
  n = c(45, 38, 52),
  report_date = as.Date("2024-01-31")
)

hist_formatted <- format_histogram(hist_simple, end_date = "2024-01-31")
print(hist_formatted)
```

# Core Analysis Functions

## Queue Size

Calculate the total number of patients waiting:

```{r queue-size}
# Queue size at end of January
queue_jan <- wl_queue_size_hist(wl_hist_jan)
cat("Queue size (31 Jan):", queue_jan, "patients\n")

# Queue size at end of February
queue_feb <- wl_queue_size_hist(wl_hist_feb)
cat("Queue size (29 Feb):", queue_feb, "patients\n")
```

## Referral Statistics (Demand)

Calculate demand statistics from the most recent snapshot:

```{r referral-stats}
referral_stats <- wl_referral_stats_hist(wl_hist_feb)
print(referral_stats)
```

This shows:
- `demand_weekly`: Average weekly referrals
- `demand_daily`: Average daily referrals
- `demand_cov`: Coefficient of variation (uncertainty in arrival pattern)
- `demand_count`: Total referrals over the period

## Removal Statistics (Capacity)

Calculate removal/capacity statistics by comparing multiple snapshots:

```{r removal-stats}
# Uses both Jan and Feb snapshots to estimate removals
removal_stats <- wl_removal_stats_hist(wl_hist)
print(removal_stats)
```

This shows:
- `capacity_weekly`: Average weekly removals (treatments)
- `capacity_daily`: Average daily removals
- `capacity_cov`: Coefficient of variation
- `removal_count`: Total removals between snapshots

The function compares consecutive snapshots to detect where counts decreased (indicating removals).

## Mean Waiting Time

Calculate average waiting time from a snapshot:

```{r mean-wait}
mean_wait <- wl_mean_wait_age_hist(wl_hist_feb)
cat("Mean waiting time:", round(mean_wait, 1), "weeks\n")
```

## Percentile Waiting Times

Find specific percentiles of the waiting time distribution:

```{r percentiles}
# Median (50th percentile)
p50 <- wl_percentile_hist(wl_hist_feb, percentage = 50)
cat("Median wait:", round(p50$weeks, 1), "weeks\n")

# 95th percentile
p95 <- wl_percentile_hist(wl_hist_feb, percentage = 95)
cat("95th percentile wait:", round(p95$weeks, 1), "weeks\n")
```

# Comprehensive Statistics with wl_stats_hist()

The `wl_stats_hist()` function provides a complete analysis in one call:

```{r comprehensive-stats}
# Calculate all key statistics
stats <- wl_stats_hist(wl_hist, target_wait = 18)
print(stats)
```

This returns a data frame with:

- **Demand/Capacity metrics**: Mean demand, mean capacity, load (demand/capacity ratio)
- **Queue metrics**: Current size, target size, whether queue is too big
- **Wait time metrics**: Mean wait, pressure indicator
- **Planning metrics**: Target capacity, relief capacity (if needed)
- **Variability**: Coefficients of variation for arrivals and removals

### Understanding the Results

```{r interpret-stats}
cat("Current queue:", stats$queue_size, "patients\n")
cat("Target queue:", round(stats$target_queue_size), "patients\n")
cat("Queue too big?", stats$queue_too_big, "\n\n")

cat("Mean demand:", round(stats$mean_demand, 1), "patients/week\n")
cat("Mean capacity:", round(stats$mean_capacity, 1), "patients/week\n")
cat("Load (demand/capacity):", round(stats$load, 2), "\n")
cat("Load too big?", stats$load_too_big, "\n\n")

cat("Mean waiting time:", round(stats$mean_wait_age, 1), "weeks\n")
cat("Target wait:", 18, "weeks\n")
cat("Pressure (2 × mean / target):", round(stats$pressure, 2), "\n")
```

**Key indicators:**

- `load >= 1.0`: System is unstable (demand exceeds capacity)
- `queue_too_big = TRUE`: Queue exceeds twice the target size
- `pressure > 1.0`: Unlikely to meet waiting time targets

# Working with Multiple Snapshots

## Time Series Analysis

```{r time-series-example}
# Create a longer time series (3 months)
wl_hist_mar <- data.frame(
  arrival_since = as.Date(c("2024-03-20", "2024-03-13", "2024-03-06", "2024-02-28")),
  arrival_before = as.Date(c("2024-03-26", "2024-03-19", "2024-03-12", "2024-03-05")),
  n = c(48, 41, 50, 58),
  report_date = as.Date("2024-03-31")
)

wl_hist_full <- rbind(wl_hist_jan, wl_hist_feb, wl_hist_mar)

# Calculate stats using all three time points
stats_full <- wl_stats_hist(wl_hist_full, target_wait = 18)

# Removal stats averaged across Jan→Feb and Feb→Mar
removal_stats_full <- wl_removal_stats_hist(wl_hist_full)
print(removal_stats_full)
```

## Filtering Date Ranges

Specify which time period to analyze:

```{r date-filters}
# Only analyze Feb→Mar period
stats_feb_mar <- wl_stats_hist(
  wl_hist_full,
  target_wait = 18,
  start_date = "2024-02-01",
  end_date = "2024-03-31"
)

cat("Capacity (Feb-Mar only):", round(stats_feb_mar$mean_capacity, 1), "\n")

# Compare to full period
cat("Capacity (Jan-Mar):", round(stats_full$mean_capacity, 1), "\n")
```

# Working with Categorical Variables

## Multiple Specialties

```{r categorical-data}
# Histogram with multiple specialties
wl_hist_specialty <- data.frame(
  specialty = rep(c("Cardiology", "Orthopedics"), each = 3),
  arrival_since = rep(as.Date(c("2024-01-24", "2024-01-17", "2024-01-10")), 2),
  arrival_before = rep(as.Date(c("2024-01-30", "2024-01-23", "2024-01-16")), 2),
  n = c(45, 38, 52, 30, 42, 28),
  report_date = as.Date("2024-01-31")
)

print(wl_hist_specialty)
```

## Aggregating Across Categories

Remove categorical splits to get overall statistics:

```{r aggregate-histogram}
# Aggregate across specialties
wl_hist_total <- aggregate_histogram(wl_hist_specialty)
print(wl_hist_total)

# Calculate overall queue size
total_queue <- wl_queue_size_hist(wl_hist_total)
cat("Total queue (all specialties):", total_queue, "\n")
```

## Keeping Categories

Analyze by specialty:

```{r by-category}
# Keep specialty grouping
wl_hist_by_specialty <- aggregate_histogram(
  wl_hist_specialty, 
  group_columns = "specialty"
)

print(wl_hist_by_specialty)
```

# Practical Example: NHS National Data Pattern

This example shows a typical workflow for NHS published statistics:

```{r nhs-example}
# Simulate NHS RTT data structure (weekly bins)
create_nhs_snapshot <- function(report_date, base_count = 100) {
  # NHS typically reports in 1-week bins up to 52+ weeks
  weeks <- c(1, 2, 3, 4, 6, 8, 10, 12, 15, 18, 21, 26, 52)
  
  data.frame(
    arrival_since = as.Date(report_date) - (weeks * 7 + 6),
    arrival_before = as.Date(report_date) - (weeks * 7),
    n = round(base_count * exp(-weeks/20)),  # Exponential decay
    report_date = as.Date(report_date)
  )
}

# Create quarterly snapshots
nhs_q4_2023 <- create_nhs_snapshot("2023-12-31", base_count = 120)
nhs_q1_2024 <- create_nhs_snapshot("2024-03-31", base_count = 110)
nhs_q2_2024 <- create_nhs_snapshot("2024-06-30", base_count = 105)

nhs_hist <- rbind(nhs_q4_2023, nhs_q1_2024, nhs_q2_2024)

# Analyze NHS data
nhs_stats <- wl_stats_hist(nhs_hist, target_wait = 18)

cat("NHS Example Analysis\n")
cat("====================\n")
cat("Queue size (Jun 2024):", nhs_stats$queue_size, "\n")
cat("Target queue size:", round(nhs_stats$target_queue_size), "\n")
cat("Mean waiting time:", round(nhs_stats$mean_wait_age, 1), "weeks\n")
cat("Meeting 18-week target?", nhs_stats$pressure <= 1.0, "\n")
```

# Comparison: Histogram vs Patient-Level

## Creating Equivalent Data

```{r comparison-setup}
# Patient-level data
set.seed(42)
referral_dates <- seq(as.Date("2024-01-01"), as.Date("2024-01-31"), by = "day")
patient_data <- data.frame(
  referral = sample(referral_dates, 100, replace = TRUE),
  removal = as.Date(NA)
)

# Simulate some removals
removal_indices <- sample(1:100, 30)
# Fix: use as.Date() to keep it as Date type
patient_data$removal[removal_indices] <- as.Date(
  patient_data$referral[removal_indices] + sample(7:60, 30, replace = TRUE),
  origin = "1970-01-01"
)

head(patient_data)
```

## Patient-Level Analysis

```{r patient-level-analysis}
# Analyze with patient-level function
patient_stats <- wl_stats(patient_data, target_wait = 18)
patient_queue <- wl_queue_size(patient_data)

cat("Patient-level approach:\n")
cat("Queue size:", patient_queue$queue_size, "\n")  # Changed: extract queue_size
cat("Mean demand:", round(patient_stats$mean_demand, 1), "per week\n")
cat("Mean capacity:", round(patient_stats$mean_capacity, 1), "per week\n")
```

## Histogram Analysis (Same Data)

```{r histogram-from-patient}
# Convert to histogram format (simplified)
# Create bins: patients still waiting, grouped by referral week
waiting_patients <- patient_data[is.na(patient_data$removal), ]

# This is a simplified example - real conversion would be more complex
hist_from_patient <- data.frame(
  arrival_since = as.Date("2024-01-01"),
  arrival_before = as.Date("2024-01-31"),
  n = nrow(waiting_patients),
  report_date = as.Date("2024-01-31")
)

hist_queue <- wl_queue_size_hist(hist_from_patient)
cat("\nHistogram approach:\n")
cat("Queue size:", hist_queue, "\n")
```

## Trade-offs

**Patient-level approach:**
- ✓ More precise calculations
- ✓ Can track individual patient journeys
- ✗ Requires detailed data
- ✗ Privacy concerns
- ✗ Slower with large datasets

**Histogram approach:**
- ✓ Works with aggregate published data
- ✓ Privacy-preserving
- ✓ Efficient for large datasets
- ✗ Less precision (binned data)
- ✗ Cannot track individuals
- ✗ Requires multiple snapshots for removals

# Summary

## Key Functions

| Function | Purpose | Input |
|----------|---------|-------|
| `format_histogram()` | Prepare/validate data | Histogram with arrival_since |
| `aggregate_histogram()` | Collapse categories | Histogram with categorical variables |
| `wl_queue_size_hist()` | Current queue size | Single snapshot |
| `wl_referral_stats_hist()` | Demand statistics | Single snapshot |
| `wl_removal_stats_hist()` | Capacity statistics | Multiple snapshots |
| `wl_mean_wait_age_hist()` | Mean waiting time | Single snapshot |
| `wl_percentile_hist()` | Percentile waits | Single snapshot |
| `wl_stats_hist()` | Complete analysis | Multiple snapshots |

## Best Practices

1. **Always include report_date**: Essential for multi-snapshot analysis
2. **Use consistent time bins**: Weekly or monthly bins work best
3. **Multiple snapshots for accuracy**: More snapshots = better capacity estimates
4. **Validate your data**: Use `format_histogram()` to ensure structure
5. **Check for missing dates**: Ensure consecutive snapshots for reliable trends

## Next Steps

- Explore the [simulation vignette](waiting_list_sim.html) for forward projections
- See [example walkthrough](example_walkthrough.html) for patient-level analysis
- Read about [queuing theory foundations](three_example_waiting_lists.html)

# References

This package implements methods described in:

Fong et al. (2022). "Queueing Theory and Machine Learning Applied to National Scale Medical Waiting Lists." *medRxiv*. doi:10.1101/2022.08.23.22279117
